---
title: "ToxCast Summaries"
author: "Steven R. Corsi, Laura A. De Cicco"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    fig_caption: yes

---



```{r message=FALSE, echo=FALSE}
library(pander)
library(dplyr)
library(reshape2)
library(DT)
library(toxEval)
library(knitr)

endPointInfo <- endPointInfo
# AC50 data provided in the toxEval package:
AC50gain <- AC50gain
packagePath <- system.file("extdata", package="toxEval")
#Site Info:
filePath <- file.path(packagePath, "stationINFO.RData")
load(file=filePath)
#parameter info:
filePath <- file.path(packagePath, "pCodeInfo.RData")
load(file=filePath)

pCodeInfo$class <- gsub("(^|[[:space:]])([[:alpha:]])", "\\1\\U\\2", tolower(pCodeInfo$class), perl=TRUE)
pCodeInfo$class[pCodeInfo$class == "Pah"] <- "PAH"
pCodeInfo$class[pCodeInfo$class == "Human Drug, Non Prescription"] <- "Pharmaceuticals"

pCodeInfo <- pCodeInfo[pCodeInfo$casrn != "", ]
packagePath <- system.file("extdata", package="toxEval")
filePath <- file.path(packagePath, "waterSamples.RData")
load(file=filePath)

#Correct DEET:
# casDEET <- AC50gain$casn[AC50gain$chnm == "DEET"]
# pCodeDEET <- pCodeInfo$parameter_cd[pCodeInfo$casrn == casDEET]
# dlDEET <- min(waterSamples$valueToUse_61947, na.rm = TRUE)
# waterSamples$qualifier_61947 <- ifelse(waterSamples$valueToUse_61947 <= dlDEET*4, "<", "")
# Just removing DEET:
waterSamples <- waterSamples[,-grep("61947",x = names(waterSamples))]

waterSamples$site["USGS-04157005" == waterSamples$site] <- "USGS-04157000"

waterSamplePCodes <- names(waterSamples)[grep("valueToUse", 
                                              names(waterSamples))]
waterSamplePCodes <- sapply(strsplit(waterSamplePCodes, "_"), function(x) x[2])


AC50 <- left_join(AC50gain, 
                  pCodeInfo[pCodeInfo$parameter_cd %in% waterSamplePCodes,
                            c("casrn", "parameter_units", "mlWt")],
                   by= c("casn"="casrn")) %>%
  filter(!is.na(parameter_units)) %>%
  rename(desiredUnits = parameter_units) %>%
  mutate(conversion = mlWt) %>%
  select(casn, chnm, desiredUnits, mlWt, conversion)

AC50Converted <- left_join(AC50, AC50gain, by = c("casn", "chnm"))
infoColumns <- c("casn", "chnm", "desiredUnits","mlWt", "conversion", "code","chid")

endPointData <- AC50Converted[,!(names(AC50Converted) %in% infoColumns)]
endPointData <- 10^endPointData
endPointData <- endPointData * AC50Converted$conversion
endPoint <- cbind(AC50, data.frame(endPointData))
endPoint <- rename(endPoint, Units=desiredUnits)

qualColumns <- grep("qualifier", names(waterSamples))
valColumns <- grep("valueToUse", names(waterSamples))

waterData <- waterSamples[,valColumns]
waterData[waterSamples[,qualColumns] == "<"] <- 0

newSiteKey <- setNames(stationINFO$shortName, stationINFO$fullSiteID)
wData <- cbind(waterSamples[,1:2],waterData)

valColumns <- grep("valueToUse",names(wData))

newSiteKey <- setNames(stationINFO$shortName, stationINFO$fullSiteID)

wData <- cbind(waterSamples[,1:2],waterData)
valColumns <- grep("valueToUse",names(wData))

chemicalSummary <- wData %>%
  melt(id.vars=c("ActivityStartDateGiven","site")) %>%
  mutate(variable=as.character(variable)) %>%
  filter(!is.na(value)) %>%
  rename(measuredValue=value, pCode=variable, date=ActivityStartDateGiven) %>%
  mutate(pCode=sapply(strsplit(pCode, "_"), function(x) x[2])) %>%
  left_join(pCodeInfo[c("parameter_cd","casrn","class","srsname")], by=c("pCode"="parameter_cd")) %>%
  select(srsname,casrn, class, measuredValue, site, date) %>%
  right_join(endPoint, by=c("casrn"="casn")) %>%
  select(-mlWt, -conversion, -casrn,  -Units, -srsname) %>%
  melt(id.vars = c("class", "site", "measuredValue","chnm","date")) %>% 
  filter(!is.na(value)) %>%
  mutate(variable=as.character(variable)) %>%
  rename(endPointValue=value, endPoint=variable) %>%
  mutate(EAR=measuredValue/endPointValue) 

chemSum1 <- chemicalSummary %>%
  group_by(chnm,  class, site, date) %>%
  summarize(hits= as.numeric(any(EAR > 0.1))) %>%
  group_by(chnm,  class, site) %>%
  summarize(hits=as.numeric(any(hits > 0))) %>%
  group_by(chnm, class) %>%
  summarize(freq=sum(hits)/length(unique(site)),
            nSites=sum(hits)) 

chemSum2 <- chemicalSummary %>%
  group_by(chnm,  class) %>%  
  summarize(maxEAR=max(EAR), 
            nSamples=nrow(unique(data.frame(date,site))),
            nEndPoints=length(unique(endPoint[EAR > 0.1]))) %>%
  left_join(chemSum1, by=c("chnm","class")) %>%
  data.frame %>%
  arrange(desc(freq)) %>%
  select(chnm, class, freq, maxEAR, nEndPoints, nSites, nSamples)



siteSummary <- wData %>%
  melt(id.vars=c("ActivityStartDateGiven","site")) %>%
  mutate(variable=as.character(variable)) %>%
  filter(!is.na(value)) %>%
  rename(measuredValue=value, pCode=variable, date=ActivityStartDateGiven) %>%
  mutate(pCode=sapply(strsplit(pCode, "_"), function(x) x[2])) %>%
  left_join(pCodeInfo[c("parameter_cd","casrn","class","srsname")], by=c("pCode"="parameter_cd")) %>%
  select(srsname,casrn, class, measuredValue, site, date) %>%
  right_join(endPoint, by=c("casrn"="casn")) %>%
  select(-mlWt, -conversion, -casrn,  -Units, -srsname) %>%
  melt(id.vars = c("class", "site", "date","measuredValue","chnm")) %>% 
  filter(!is.na(value)) %>%
  mutate(variable=as.character(variable)) %>%
  rename(endPointValue=value, endPoint=variable) %>%
  mutate(EAR=measuredValue/endPointValue) %>%
  mutate(site=newSiteKey[site]) %>%
  select(site, chnm, EAR, endPoint, class, date) %>% 
  arrange(site, chnm, EAR)

summ1 <- siteSummary %>%
  group_by(site, date) %>%
  summarize(hits= as.numeric(any(EAR > 0.1))) %>%
  group_by(site) %>%
  summarize(freq=sum(hits)/length(unique(date)),
            nSamples=length(unique(date))) 

summ2 <-  siteSummary %>%
  group_by(site) %>%
  summarize(nChem = length(unique(chnm[EAR > 0.1])),
            nEndPoints = length(unique(endPoint[EAR > 0.1])),
            maxEAR = max(EAR))


summary <- left_join(summ1, summ2, by="site") %>%
  arrange(desc(maxEAR))


```


# Water Sample Data
The data returned from water sample tests was returned in a 'wide' format. The first job is to find the unique units. This data is arranged by a 5 digit USGS parameter code. Using the **dataRetrieval** package, information about the parameter codes can be retrieved. Each parameter code has 6 columns: detectionLimit, rep_lev_cd, qualifier, valueToUse, val_qual_tex, meth_cd, dqi_cd. We can pull out one of those columns to find the unique paremeter codes, then use **webchem** to get the molecular weights. In this dataset, all the data is reported in `r unique(pCodeInfo$parameter_units)`. Our conversion therefore is:

$$
\require{cancel}
\begin{aligned}
Reported Data = \log \mu M \\
\mu M = 10^{\log \mu M} \\
MW = \frac{g}{mol} , M = \frac{mol}{L}  \\
\cancel{\mu M} \left[\frac{\cancel{M}}{\cancel{10^6} \cancel{\mu M}}\right] \left[\frac{1}{\cancel{M}}  \frac{\cancel{mol}}{L}\right] \left[ \frac{MW \cancel{g}}{\cancel{mol}}\right] \left[\frac{\cancel{10^6} \mu g}{\cancel{g}}\right] = MW * \frac{\mu g}{L} 
\end{aligned}
$$

The next task is to determine if the detectionLimit is going to tell us anything in relation to the ToxCast endpoints. 


## Water Sample Detection Levels

For now, we'll focus on the minimum detection level:
```{r warning=FALSE, echo=FALSE}
pCodeSummary <- select(pCodeInfo, srsname, casrn, parameter_units, mlWt) %>%
  filter(pCodeInfo$parameter_cd %in% waterSamplePCodes)

detectionLimits <- waterSamples[,grep("detectionLimit", names(waterSamples))]
detectionLimits <- na.omit(detectionLimits)

detLevels  <- data.frame(dl_pcode=names(detectionLimits),
                         minDetLevel=apply(detectionLimits, 2, min),
                         row.names=NULL, stringsAsFactors=FALSE) %>%
  mutate(pCode=sapply(strsplit(dl_pcode, "_"), function(x) x[2])) %>%
  left_join(pCodeInfo[c("parameter_cd","casrn","class")], 
            by=c("pCode"="parameter_cd")) %>%
  select(casrn, class, minDetLevel) %>%
  right_join(endPoint, by=c("casrn"="casn")) %>%
  select(-mlWt, -conversion, -casrn,  -Units) %>%
  melt(id.vars = c("class", "minDetLevel","chnm")) %>% 
  filter(!is.na(value)) %>%
  mutate(variable=as.character(variable)) %>%
  rename(endPointValue=value, endPoint=variable) %>%
  mutate(EAR=minDetLevel/endPointValue) %>%
  filter(EAR >= 0.1) %>%
  select(Chemical=chnm, Class=class, EAR, endPoint, minDetLevel) %>%
  arrange(desc(EAR)) 


datatable(detLevels, rownames = FALSE,options = list(pageLength = 25))  %>% 
    formatRound(c("EAR","minDetLevel"), digits = 1)

```

This table shows us the chemicals that `r detLevels$chnm` may lead to issues since the detection levels are higher than some of the end point values.

In this case therefore, we cannot simply ignore all of the left-censored values. However, for the time being, we will use only non-censored values in our analysis. Let's begin by trimming down the waterSamples data frame to only the relavent information. Here, we are setting all left-censored values to NA, we could instead use the detection limit.


## Chemical Summary Tables: Water Samples

### Overall
```{r, echo=FALSE, results='asis'}
  
datatable(chemSum2, rownames = FALSE, 
            options = list(pageLength = 10), 
            colnames=c("Chemical","Class","Fraction of Hits Per Site","Maximum EAR","Number End Points", "Sites With Hits", "Total Samples")) %>% 
    formatRound(c("maxEAR", "freq"), digits = 3)
```

### Individual EndPoints
Looking at individual endpoints (filtering only rows that have some hit):

```{r message=FALSE, results='asis', echo=FALSE}


chemSum3 <- chemicalSummary %>%
  group_by(chnm,  class, site, date, endPoint) %>%
  summarize(hits= as.numeric(any(EAR > 0.1))) %>%
  group_by(chnm,  class, site, endPoint) %>%
  summarize(hits=as.numeric(any(hits > 0))) %>%
  group_by(chnm, class, endPoint) %>%
  summarize(freq=sum(hits)/length(unique(site)),
            nSites=sum(hits)) 

chemSum4 <- chemicalSummary %>%
  group_by(chnm,  class, endPoint) %>%  
  summarize(maxEAR=max(EAR), 
            nSamples=nrow(unique(data.frame(date,site)))) %>%
  left_join(chemSum3, by=c("chnm","class","endPoint")) %>%
  select(chnm, class, freq, maxEAR, nSites, nSamples, endPoint) %>%
  left_join(endPointInfo, by=c("endPoint"="assay_component_endpoint_name")) %>% 
  select(chnm, maxEAR, freq, nSites, 
         endPoint, class, contains("intended_target_")) %>%
  rename(type=intended_target_type, type_sub=intended_target_type_sub,
         family=intended_target_family, family_sub=intended_target_family_sub) %>%
  select(-contains("intended_target_"))  %>%
  data.frame %>%
  arrange(desc(freq)) 


  datatable(unique(chemSum4), rownames = FALSE, 
            options = list(pageLength = 10)) %>% 
    formatRound(c("maxEAR", "freq"), digits = 2)

```

## Site Summaries: Water Samples

The first site summary table groups only by site:

```{r, echo=FALSE, results='asis'}
summary_display <- left_join(summary, stationINFO[,c("shortName","Ag.pct","Urban.pct")], by=c("site"="shortName")) 

datatable(summary_display, rownames = FALSE, 
          colnames=c("Station Name","Fraction of Hits Per Samples", "Total Samples","Number of Chemicals With Hits", "Number End Points", "Maximum EAR","Ag [%]","Urban [%]")) %>% 
  formatRound(c("freq","maxEAR","Ag.pct","Urban.pct"), digits = 2)
```

## ToxCast: Water Samples Sites Boxplots

```{r fig.width=7, fig.height=7, echo=FALSE,warning=FALSE, message=FALSE}
library(ggplot2)
library(grid)
summary_display1 <- transform(summary_display, 
                             site = reorder(site, order(nChem,decreasing=TRUE)))

siteLimits <- filter(summary_display, freq>0) %>%
  arrange(desc(nChem)) %>%
  select(site,nEndPoints,nChem)


summary_display2 <- siteSummary %>%
  filter(EAR>0.1)


ggplot(summary_display2, aes(x=site, y=EAR)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.25), 
        legend.position = "none")+
  ylim(-2,50) + 
  scale_x_discrete(limits=siteLimits$site)+
  labs(x = "Site (ordered by number of chemicals with hits)") +
  geom_text(data=data.frame(), aes(x=siteLimits$site, y=-1.5,
                                   label=siteLimits$nEndPoints),size=3) 
```

## ToxCast: Water Samples Chemical Boxplots

```{r fig.width=7, fig.height=7, echo=FALSE,warning=FALSE, message=FALSE}

chemLimits <- filter(chemSum2, freq>0) %>%
  arrange(desc(freq)) %>%
  select(chnm, nEndPoints)

chemicalSummary_display <- chemicalSummary %>%
  filter(EAR>=0.1) 

ggplot(chemicalSummary_display, aes(x=chnm, y=EAR)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.25), 
        legend.position = "none") +
  ylim(-2,50) + 
  scale_x_discrete(limits=chemLimits$chnm) +
  labs(x = "Chemical Name (ordered by fraction of sites with hits)")+
  geom_text(data=data.frame(), aes(x=chemLimits$chnm, y=-1.5,
                                   label=chemLimits$nEndPoints),size=3) 

```


## ToxCast: Water Samples Map
<b>Color:</b> Maximum EAR<br/>
<b>Size:</b> Number of chemicals with hits (0-19)

```{r, echo=FALSE, results='asis', echo=FALSE, fig.width=7, fig.height=5}

library(leaflet)
stationINFO <- stationINFO[!is.na(stationINFO$dec.lat.va),]

mapData <- left_join(stationINFO, summary, by=c("shortName"="site"))
mapData <- mapData[!is.na(mapData$dec.lat.va),]
mapData <- mapData[!is.na(mapData$nChem),]


col_types <- c("darkblue","dodgerblue","green","yellow","orange","red")
leg_vals <- c(0,1,5,10,100,1000,2500)#seq(0,2500, 500) # This is maxEAR 

cols <- colorNumeric(col_types, domain = leg_vals)
rad <- seq(5000,15000, 500)
mapData$sizes <- rad[as.numeric(cut(mapData$nChem, c(-1:19)))]

pal = colorBin(col_types, mapData$maxEAR, bins = leg_vals)

m = leaflet(mapData) %>%
  addProviderTiles("Esri.WorldPhysical") %>%
  setView(lng = -83.5, lat = 44.5, zoom=6)%>%
  addCircles(lat=~dec.lat.va, lng=~dec.long.va, 
             popup=paste('<b>',mapData$Station.Name,'</b><br/>',
                         "<table><tr><td>Frequency</td><td>",sprintf("%1.1f", mapData$freq),'</td></tr>',
                         "<tr><td>nSample</td><td>",mapData$nSamples,'</td></tr>',
                         "<tr><td>maxEAR</td><td>",sprintf("%1.1f", mapData$maxEAR),'</td></tr>',
                         "<tr><td>nChem</td><td>",mapData$nChem,'</td></tr>',
                         "<tr><td>nEndPoints</td><td>", mapData$nEndPoints,'</td></tr></table>'),
              fillColor = ~pal(maxEAR), 
             weight = 1,
             color = "black",
             fillOpacity = 0.8, radius = ~sizes, opacity = 0.8) %>%
    addLegend(
      position = 'bottomleft',
      pal=pal,
      values=~maxEAR,
      opacity = 0.8,
      title = 'Maximum EAR') 

m

```

# ToxCast: Passive Data

```{r, echo=FALSE}

packagePath <- system.file("extdata", package="toxEval")
filePath <- file.path(packagePath, "passiveData.RData")
load(file=filePath)
# #Site Info:
# filePath <- file.path(packagePath, "stationINFO.RData")
# load(file=filePath)
# #parameter info:
# filePath <- file.path(packagePath, "pCodeInfo.RData")
# load(file=filePath)

#passive class info:
passiveCAS <- readRDS(file.path(packagePath, "passiveCAS.rds"))

pCodeInfo <- pCodeInfo[pCodeInfo$casrn != "", ]

# Some chemicals have 2 rows (different units)...
# To simplify for now, we'll ignore the duplicate chemicals:


passiveData <- left_join(passiveData, passiveCAS, by=c("CAS", "Chemical"))
passiveData <- passiveData[!duplicated(passiveData$CAS),]
passiveData$class[passiveData$Chemical == "Hexachlorobenzene (HCB)"] <-"OC Pesticides"
passiveData$class[passiveData$Chemical == "Pentachloroanisole (PCA)"] <-"OC Pesticides"
passiveData$class[is.na(passiveData$class)] <-"Pharmaceuticals"

passiveData <- passiveData[,-which(names(passiveData) == "site04249000.1")]


unitConversion <- setNames(c(10^6, 10^3), c("pg/L", "ng/L") )

AC50 <- right_join(AC50gain[,c("casn"), drop=FALSE], 
                   passiveData[,c("CAS", "Units", "mlWt","Chemical")],
                   by= c("casn" = "CAS")) %>%
  rename(desiredUnits = Units, chnm=Chemical) %>%
  filter(!is.na(mlWt)) %>%
  mutate(conversion = unitConversion[desiredUnits] * mlWt) %>%
  select(casn, chnm, desiredUnits, mlWt, conversion)

AC50Converted <- left_join(AC50, AC50gain, by = c("casn", "chnm"))
infoColumns <- c("casn", "chnm", "desiredUnits","mlWt", "conversion", "code","chid")

endPointData <- AC50Converted[,!(names(AC50Converted) %in% infoColumns)]
endPointData <- 10^endPointData
endPointData <- endPointData * AC50Converted$conversion
endPoint <- cbind(AC50, data.frame(endPointData))
endPoint <- rename(endPoint, Units=desiredUnits)

siteColumns <- grep("site",names(passiveData))
passiveData[,siteColumns] <- suppressWarnings(
  sapply(passiveData[,siteColumns], function(x) as.numeric(x)))

#For this analysis, we'll consider NA's to be 0 (other options exist):
passiveData[,siteColumns][is.na(passiveData[,siteColumns])] <- 0

newSiteKey <- setNames(stationINFO$shortName, stationINFO$fullSiteID)

infoColumns <- c("Chemical", "CAS", "Units", "MLD", "MQL", "mlWt","class")

chemicalSummary_passive <- melt(passiveData, id.vars = infoColumns) %>%
  mutate(variable=as.character(variable)) %>%
  filter(!is.na(value)) %>%
  select(srsname=Chemical,
         casrn=CAS,
         class=class,
         measuredValue=value, 
         site=variable) %>%
  right_join(endPoint, by=c("casrn"="casn")) %>%
  select(-mlWt, -conversion, -casrn,  -Units, -srsname) %>%
  melt(id.vars = c("class", "site", "measuredValue","chnm")) %>% 
  filter(!is.na(value)) %>%
  mutate(variable=as.character(variable)) %>%
  rename(endPointValue=value, endPoint=variable) %>%
  mutate(EAR=measuredValue/endPointValue) 

chemSum1_passive <- chemicalSummary_passive %>%
  group_by(chnm,  class, site) %>%
  summarize(hits= as.numeric(any(EAR > 0.1))) %>%
#   group_by(chnm,  class, site) %>%
#   summarize(hits=as.numeric(any(hits > 0))) %>%
  group_by(chnm, class) %>%
  summarize(freq=sum(hits)/length(unique(site)),
            nSites=sum(hits)) 

chemSum2_passive <- chemicalSummary_passive %>%
  group_by(chnm,  class) %>%  
  summarize(maxEAR=max(EAR), 
            nSamples=1,
            nEndPoints=length(unique(endPoint[EAR > 0.1]))) %>%
  left_join(chemSum1_passive, by=c("chnm","class")) %>%
  data.frame %>%
  select(chnm, class, freq, maxEAR, nEndPoints, nSites, nSamples) %>%
  arrange(desc(freq)) 

siteSummary_passive <- melt(passiveData, id.vars = infoColumns) %>%
  mutate(variable=as.character(variable)) %>%
  filter(!is.na(value)) %>%
  select(srsname=Chemical,
         casrn=CAS,
         class=class,
         measuredValue=value, 
         site=variable) %>%
  right_join(endPoint, by=c("casrn"="casn")) %>%
  select(-mlWt, -conversion, -casrn,  -Units, -srsname) %>%
  melt(id.vars = c("class", "site", "measuredValue","chnm")) %>% 
  filter(!is.na(value)) %>%
  mutate(variable=as.character(variable)) %>%
  rename(endPointValue=value, endPoint=variable) %>%
  mutate(EAR=measuredValue/endPointValue) %>%
  mutate(site=newSiteKey[paste0("USGS-",gsub("[^0-9]", "", site))]) %>%
  select(site, chnm, EAR, endPoint, class) %>% 
  arrange(site, chnm, EAR)

summ1_passive <- siteSummary_passive %>%
  group_by(site) %>%
  summarize(hits= as.numeric(any(EAR > 0.1)),
            nChem = length(unique(chnm[EAR > 0.1])),
            nEndPoints = length(unique(endPoint[EAR > 0.1])),
            maxEAR = max(EAR))%>%
  arrange(desc(maxEAR))


```

## Chemical Summary Tables: Water Samples

### Overall

```{r, echo=FALSE, results='asis'}
chemSum2_passive <- chemSum2_passive[,-7] %>%
  filter(freq > 0)

datatable(chemSum2_passive, rownames = FALSE, 
            options = list(pageLength = 10),
            colnames=c("Chemical","Class","Fraction of Hits Per Site","Maximum EAR","Number of End Points", "Sites With Hits")) %>% 
    formatRound(c("maxEAR", "freq"), digits = 4)

```


### Individual EndPoints
Looking at individual endpoints (filtering only rows that have some hit):

```{r message=FALSE, results='asis', echo=FALSE}


chemSum3_passive <- chemicalSummary_passive %>%
  group_by(chnm,  class, site, endPoint) %>%
  summarize(hits= as.numeric(any(EAR > 0.1))) %>% 
  group_by(chnm, class, endPoint) %>%
  summarize(freq=sum(hits)/length(unique(site)),
            nSites=sum(hits)) 

chemSum4_passive <- chemicalSummary_passive %>%
  group_by(chnm,  class, endPoint) %>%  
  summarize(maxEAR=max(EAR), 
            nSamples=1) %>%
  left_join(chemSum3_passive, by=c("chnm","class","endPoint")) %>%
  select(chnm, class, freq, maxEAR, nSites, nSamples, endPoint) %>%
  left_join(endPointInfo, by=c("endPoint"="assay_component_endpoint_name")) %>% 
  select(chnm, maxEAR, freq, nSites, 
         endPoint, class, contains("intended_target_")) %>%
  rename(type=intended_target_type, type_sub=intended_target_type_sub,
         family=intended_target_family, family_sub=intended_target_family_sub) %>%
  select(-contains("intended_target_"))  %>%
  data.frame %>%
  arrange(desc(freq))  %>%
  filter(freq > 0)


  datatable(unique(chemSum4_passive), rownames = FALSE, 
            options = list(pageLength = 10)) %>% 
    formatRound(c("maxEAR", "freq"), digits = 2)

```

## Site Summaries: Passive Data

The first site summary table groups only by site:

```{r, echo=FALSE, results='asis'}
summary_display_passive <- left_join(summ1_passive, stationINFO[,c("shortName","Ag.pct","Urban.pct")], by=c("site"="shortName")) 
  
summary_display_passive <- summary_display_passive[,-2]
datatable(summary_display_passive, rownames = FALSE,
          colnames=c("Station", "Number of Chemicals with Hits", "Number End Points", "Maximum EAR", "Ag [%]", "Urban [%]")) %>% 
#           options = list(pageLength = 50)) %>% 
  formatRound(c("maxEAR"), digits = 2)
```


## ToxCast: Passive Data Map 
<b>Color:</b> Maximum EAR<br/>
<b>Size:</b> Number of chemicals with hits (0-3)

```{r, echo=FALSE, results='asis', echo=FALSE, fig.width=7, fig.height=5}

stationINFO <- stationINFO[!is.na(stationINFO$dec.lat.va),]

mapDataPassive <- left_join(stationINFO, summ1_passive, by=c("shortName"="site"))
mapDataPassive <- mapDataPassive[!is.na(mapDataPassive$dec.lat.va),]
mapDataPassive <- mapDataPassive[!is.na(mapDataPassive$maxEAR),]

col_types <- c("darkblue","dodgerblue","green","yellow","orange","red")
leg_vals <- c(0,1,2,3,5,7,9)#seq(0,2500, 500) # This is maxEAR scale

cols <- colorNumeric(col_types, domain = leg_vals)
rad <- seq(2500,15000, 2500)
mapDataPassive$sizes <- rad[as.numeric(cut(mapDataPassive$nChem, c(-1,0,1,2,3)))]
mapDataPassive$sizes[is.na(mapDataPassive$sizes)] <- 1000

pal = colorBin(col_types, mapDataPassive$maxEAR, bins = leg_vals)


map2 = leaflet(mapDataPassive) %>%
#   addProviderTiles("Hydda.Full") %>% # There are more maybe worth looking at
  addProviderTiles("Esri.WorldPhysical") %>%
  setView(lng = -83.5, lat = 44.5, zoom=6)%>%
  addCircles(lat=~dec.lat.va, lng=~dec.long.va, 
             popup=paste("<b>",mapDataPassive$Station.Name,'</b><br/>',
                         "<table><tr><td>maxEAR</td><td>",sprintf("%1.1f", mapDataPassive$maxEAR),'</td></tr>',
                         "<tr><td>nChem</td><td>",mapDataPassive$nChem,'</td></tr>',
                         "<tr><td>nEndPoints</td><td>", mapDataPassive$nEndPoints,'</td></tr></table>'),
             fillColor = ~pal(maxEAR), 
             weight = 1,
             color = "blue",
             fillOpacity = 0.8, radius = ~sizes, opacity = 0.8) %>%
    addLegend(
      position = 'bottomleft',
      pal=pal,
      values=~maxEAR,
#       colors = cols(leg_vals),
      opacity = 0.8,
#       labels = paste0(leg_vals,'-'),
      title = 'Maximum EAR')
map2

```

## ToxCast: Passive Data Sites Boxplots

```{r fig.width=7, fig.height=7, echo=FALSE,warning=FALSE, message=FALSE}

summary_display1_passive <- transform(summary_display_passive, 
                             site = reorder(site, order(nChem,decreasing=TRUE)))

siteLimits_passive <- filter(summary_display_passive, nChem>0) %>%
  arrange(desc(nChem)) %>%
  select(site,nEndPoints,nChem)


summary_display2_passive <- siteSummary_passive %>%
  filter(EAR>0.1)


ggplot(summary_display2_passive, aes(x=site, y=EAR)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.25), 
        legend.position = "none")+
  ylim(-2,10) + 
  scale_x_discrete(limits=siteLimits_passive$site)+
  labs(x = "Site (ordered by number of chemicals with hits)") +
  geom_text(data=data.frame(), aes(x=siteLimits_passive$site,
                                   y=-1.5, label=siteLimits_passive$nEndPoints),size=3) 
```


## ToxCast: Passive Chemical Boxplots

```{r fig.width=7, fig.height=7, echo=FALSE,warning=FALSE, message=FALSE}

chemLimits_passive <- filter(chemSum2_passive, freq>0) %>%
  arrange(desc(freq)) %>%
  select(chnm, nEndPoints)

chemicalSummary_display_passive <- chemicalSummary_passive %>%
  filter(EAR>=0.1) 

ggplot(chemicalSummary_display_passive, aes(x=chnm, y=EAR)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.25), 
        legend.position = "none") +
  ylim(-2,10) + 
  scale_x_discrete(limits=chemLimits_passive$chnm) +
  labs(x = "Chemical Name (ordered by fraction of sites with hits)")+
  geom_text(data=data.frame(), aes(x=chemLimits_passive$chnm, y=-1.5,
                                   label=chemLimits_passive$nEndPoints),size=3) 

```




