---
title: "Basic Workflow"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: false
    fig_height: 7
    fig_width: 7
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Basic Workflow}
  \usepackage[utf8]{inputenc}
---

# Load Data

Once the data is prepared as described in the  [Prepare Data vignette]("PrepareData.html"), it can be loaded into R using the `create_toxEval` function.

```{r startup, message=FALSE}
library(toxEval)
path_to_tox <-  system.file("extdata", package="toxEval")
file_name <- "OWC_data_fromSup.xlsx"
full_path <- file.path(path_to_tox, file_name)

tox_list <- create_toxEval(full_path)

```

The `create_toxEval` function will check that required columns are included, and give messages if there are unusual issues such as sites that don't match up with data, or chemicals that don't match up with chemical information.

# Setup Analysis

The next step for exploring the data in `toxEval` is to create a data frame that includes the EAR calculations for each measurement. This data frame (`chemicalSummary`) is required for most functions within `toxEval`. Since one chemical may have many more than one "endPoint", this data frame may be much larger then the "Data" tab in the Excel file. The general workflow for creating this data frame is:

```{r chemicalSummary}
ACClong <- get_ACC(tox_list$chem_info$CAS)
ACClong <- remove_flags(ACClong)

cleaned_ep <- clean_endPoint_info(endPointInfo)
filtered_ep <- filter_groups(cleaned_ep)

chemicalSummary <- get_chemical_summary(tox_list, ACClong, filtered_ep)
```

What's going on with this workflow?

## get_ACC

The `get_ACC` function will retrieve the activity concentration at cutoff (`ACC`) values for the specified chemicals. The data was originally downloaded for `toxEval` on October 2015 from ToxCast. The data were combined from files in the "INVITRODB_V2_LEVEL5" folder. At the time of the toxEval package release, this information was found
[here](https://www.epa.gov/chemical-research/toxicity-forecaster-toxcasttm-data) in the "ToxCast & Tox21 Data Spreadsheet" data set. 
The function `get_ACC` will convert the ACC values the ToxCast database from units of log(uM) to units of ug/L, and reformat the data to a format that can be used in `toxEval`.

## remove_flags

Through the ToxCast program quality assurance procedures, information is examined and at times, it is necessary to assign a data quality flag to a specific chemical:assay result. A `toxEval` user may want to include or exclude assay results with certain flags depending on the objectives of a given study. Assay results with specific data quality flags assigned to them can be removed based on their designated flag with the `remove_flags` function. The flags included in ToxCast, and the associated `flagsShort` value (used in the `remove_flags` function) are as follows:

| Flags   |  flagsShort |
|----------|------:|
| Borderline active |  Borderline |
| Only highest conc above baseline, active |   OnlyHighest |
| Only one conc above baseline, active | OneAbove |
| Noisy data | Noisy |
| Hit-call potentially confounded by overfitting | HitCall |
| Gain AC50 < lowest conc & loss AC50 < mean conc  | GainAC50 |
| Biochemical assay with < 50% efficacy | Biochemical |

The function is written to exclude results with the flags "Borderline", "OnlyHighest", "GainAC50", "Biochemical" as a default. If the user prefers a different list, the full list must be specified (default values will be ignored).

## clean_endPoint_info

The `clean_endPoint_info` function cleans up the `endPointInfo` table from ToxCast by filtering assays considered in the analysis based on [Blackwell](https://pubs.acs.org/doi/10.1021/acs.est.7b01613). Specifically, this function hard-codes in the removal of endPoints that are ATG sources with signal loss, and NVS with signal gain. Also, this function adds some additional categories to intended_target_family and intended_target_family_sub as described in the paper linked above.

## filter_groups

<table>
<tr><td>
This function takes the data frame from 'endPointInfo' and filters the endpoints in 3 steps. First, the user specifies the "groupCol" argument, which is a column header from `endPointInfo`. Second, the user specifies the assays to use.  Finally, the user can also choose to remove specific group(s) from the category. The default is to remove "Background Measurement" and "Undefined", but a different list of priorities may need to be used based on individual study objectives.

The default category ("groupCol") is "intended_target_family". Depending on the study, other categories may be more relevant. To discover options, examine the column names in the data frame `endPointInfo`:

```{r eval=FALSE}
names(endPointInfo)
```

By default, the BioSeek set of assays are removed. The list of assays and their short names are in the Assay Name table to the right.

</td><td>

| Assay Name | Short Name |
|----------|------:|
|Apredica|APR|
|Attagene|ATG|
|BioSeek|BSK|
|NovaScreen|NVS|
|Odyssey Thera|OT|
|Toxicity Testing|TOX21|
|CEETOX|CEETOX|
|CLD|CLD|
|TANGUAY|TANGUAY|
|NHEERL_PADILLA|NHEERL_PADILLA|
|NCCT_SIMMONS|NCCT_SIMMONS|
|ACEA Biosciences|ACEA|

</td></tr> </table>

## get_chemical_summary

The function `get_chemical_summary` takes the measured user data from the output of `create_toxEval`, and joins the data with the endPoint information provided by ToxCast, as filtered by the functions described above.

## Custom configuration{#custom_config}

Let's say we want to trim some names to make the labels more readable in graphs. Here, we clean up some chemical class labels:

```{r clean}
#Trim some names:
levels(chemicalSummary$Class)[levels(chemicalSummary$Class) == "Antimicrobial Disinfectants"] <- "Antimicrobial"
levels(chemicalSummary$Class)[levels(chemicalSummary$Class) == "Detergent Metabolites"] <- "Detergent"
levels(chemicalSummary$Class)[levels(chemicalSummary$Class) == "Flavors and Fragrances"] <- "Flavor/Fragrance"
```

And here, we convert the site lists to factors. This will make any site graph in an order that we define. Here, we take Great Lakes sites and make the order up-to-down stream:

```{r sites}
#Ordering the sites to flow "downstream" of the Great Lakes:
sitesOrdered <- c("StLouis","Nemadji","WhiteWI","Bad","Montreal",
  "PresqueIsle","Ontonagon","Sturgeon","Tahquamenon","Burns",
  "IndianaHC","StJoseph","PawPaw","Kalamazoo","GrandMI",
  "Milwaukee","Muskegon","WhiteMI","PereMarquette","Manitowoc",
  "Manistee","Fox","Oconto","Peshtigo","Menominee",
  "Indian","Cheboygan","Ford","Escanaba","Manistique",
  "ThunderBay","AuSable","Rifle","Saginaw","BlackMI",
  "Clinton","Rouge","HuronMI","Raisin","Maumee",
  "Portage","Sandusky","HuronOH","Vermilion","BlackOH",
  "Rocky","Cuyahoga","GrandOH","Cattaraugus","Tonawanda",
  "Genesee","Oswego","BlackNY","Oswegatchie","Grass",
  "Raquette","StRegis")

 tox_list$chem_site$`Short Name` <- factor(tox_list$chem_site$`Short Name`,
               levels = sitesOrdered)

lakes_ordered <- c("Lake Superior",
                  "Lake Michigan",
                  "Lake Huron",
                  "Lake Erie",
                  "Lake Ontario")

tox_list$chem_site$site_grouping <- factor(tox_list$chem_site$site_grouping,
               levels=lakes_ordered)

```

# Visualize Data

## plot_tox_boxplots {#plot_tox_boxplots}

The `plot_tox_boxplots` function creates a set of boxplots based on the original input data modified by the processing steps above, and the choice of several input options. See ["Summarizing the data"](Introduction.html#summarize_data) for a description on how the EAR values are aggregated and summarized. Choosing "Chemical Class" in the `category` argument will generate separate boxplots for each unique class. "Chemical" will generate boxplots for each individual chemical, and "Biological" will generate boxplots for each group in the selected ToxCast annotation.

```{r boxplots1, warning=FALSE, message=FALSE}

bio_box <- plot_tox_boxplots(chemicalSummary, "Biological")

# The graph can be plotted without these additional lines,
# but they allow the labels to look nicer:
gb <- ggplot2::ggplot_build(bio_box)
gt <- ggplot2::ggplot_gtable(gb)
gt$layout$clip[gt$layout$name=="panel"] <- "off"
grid::grid.draw(gt)
# Other options:
# plot_tox_boxplots(chemicalSummary, "Chemical Class")
# plot_tox_boxplots(chemicalSummary, "Chemical") 
```

The graph shows a slightly different result for a single site. First, let's set up a subset of data that we will use throughout this document to show a single site. We'll use the Maumee River data.

```{r filtersiteBox, message=FALSE, warning=FALSE}
library(dplyr)

maumee <- filter(chemicalSummary, shortName == "Maumee")
maumee_site <- filter(tox_list$chem_site, `Short Name` == "Maumee")
```

Now, we'll run `plot_tox_boxplots` on that single site. The difference in the output is that instead of listing the number of sites, it lists the number of unique chemical/endpoint combinations used to create the box plot. We've also included
the code to add a custom title to the output:

```{r maumeePlot, message=FALSE, warning=FALSE}
library(ggplot2)

maumee_plot <- plot_tox_boxplots(maumee, "Biological",title = maumee_site$Fullname[1])

gb <- ggplot2::ggplot_build(maumee_plot)
gt <- ggplot2::ggplot_gtable(gb)
gt$layout$clip[gt$layout$name=="panel"] <- "off"
grid::grid.draw(gt)

```

## plot_tox_stacks {#plot_tox_stacks}

The `plot_tox_stacks` function creates a set of stacked bar charts based on the original input data modified by the processing steps above, and the choice of several input options. See ["Summarizing the data"](Introduction.html#summarize_data) for a description on how the EAR values are aggregated and summarized. Choosing "Chemical Class" in the `category` argument will generate separate stacked bars for each unique class. "Chemical" will generate stacked bars for each individual chemical, and "Biological" will generate stacked bars for each group in the selected ToxCast annotation.


```{r stackplots1, warning=FALSE, fig.width=10}
stack_plot <- plot_tox_stacks(chemicalSummary, tox_list$chem_site, "Biological")

gb <- ggplot2::ggplot_build(stack_plot)
gt <- ggplot2::ggplot_gtable(gb)
gt$layout$clip[gt$layout$name=="panel-1-1"] = "off"
grid::grid.draw(gt)

# More options:
# plot_tox_stacks(chemicalSummary, tox_list$chem_site, "Chemical Class")
# plot_tox_stacks(chemicalSummary, tox_list$chem_site, "Chemical", include_legend = FALSE) 
```

The graph shows a slightly different result for a single site. The graph now shows each individual sample. The EAR values are no longer the mean or max at the site, but rather the sums of the individual EAR values. 

```{r siteStacks, message=FALSE, warning=FALSE, fig.width=10}
maumee_plot_stack <- plot_tox_stacks(maumee, maumee_site,"Biological", title = maumee_site$Fullname[1])

gb <- ggplot2::ggplot_build(maumee_plot_stack)
gt <- ggplot2::ggplot_gtable(gb)
gt$layout$clip[gt$layout$name=="panel"] <- "off"
grid::grid.draw(gt)

```


## plot_tox_heatmap {#plot_tox_heatmap}

The `plot_tox_heatmap` function creates heat (tile) map with sites on the x-axis, a specified grouping on the y-axis (defined by the `category` argument), and color shading defining the mean or max EAR. See ["Summarizing the data"](Introduction.html#summarize_data) for a description on how the EAR values are aggregated and summarized. The y-axis grouping can be "Biological","Chemical Class", or "Chemical". The breaks for the color scale can be customized with the `breaks` argument.

This is a function where it may be ideal to create a custom order to the sites (for example, west-to-east). See the above section ["Custom configuration"](#custom_config) for instructions on how to convert the character vector `sites` to a factor with ordered levels.

```{r heat, warning=FALSE, fig.width=10}
plot_tox_heatmap(chemicalSummary, 
                  tox_list$chem_site, 
                  category = "Biological")
# More options:
# plot_tox_heatmap(chemicalSummary, tox_list$chem_site, category = "Chemical Class")
# plot_tox_heatmap(chemicalSummary, tox_list$chem_site, category = "Chemical")
```

## plot_tox_endpoints {#plot_tox_endpoints}

The `plot_tox_endpoints` function creates a set of boxplots for each endPoint based on a selected filter. The data is first filtered down to what group is specified in the `filterBy` argument. The `filterBy` argument must match one of the unique options in the `category`. So for example, if the `category` is "Chemical Class", then the `filterBy` argument must be one of the defined "Chemical Class" options such as "Herbicide". 

After the data is filtered, a boxplot is generated for each endPoint. The EAR values that are used to create the boxplots are still the mean or max (as defined by `mean_logic`) of each site as described in ["Summarizing the data"](Introduction.html#summarize_data).

```{r endpoints, warning=FALSE}
ep_plot <- plot_tox_endpoints(chemicalSummary, 
                              category = "Biological", 
                              filterBy = "Cell Cycle")

gb <- ggplot2::ggplot_build(ep_plot)
gt <- ggplot2::ggplot_gtable(gb)
gt$layout$clip[gt$layout$name=="panel"] <- "off"
grid::grid.draw(gt)

# More options:
# plot_tox_endpoints(chemicalSummary,   
#                    category = "Chemical Class", 
#                    filterBy = "PAHs")
# plot_tox_endpoints(chemicalSummary, 
#                    category = "Chemical", 
#                    filterBy = "Atrazine")
```

# Tables

## table_tox_rank {#table_tox_rank}

```{r table_tox_rank, warning=FALSE}
library(DT)
options(DT.options = list(pageLength = 5))

table_tox_rank(chemicalSummary, category = "Biological")
# More options:
# table_tox_rank(chemicalSummary, category = "Chemical Class")
# table_tox_rank(chemicalSummary, category = "Chemical")
```

The tables show slightly different results for a single site:

```{r table_tox_rank_site, warning=FALSE}
table_tox_rank(maumee, category = "Biological")
```

## table_tox_sum {#table_tox_sum}

```{r table_tox_sum, warning=FALSE}
table_tox_sum(chemicalSummary, category = "Biological")
# More options:
# table_tox_sum(chemicalSummary, category = "Chemical Class")
# table_tox_sum(chemicalSummary, category = "Chemical")
```

The tables show slightly different results for a single site:

```{r table_tox_sum_site, warning=FALSE}
table_tox_sum(maumee, category = "Biological")
```


## table_endpoint_hits {#table_endpoint_hits}

```{r table_endpoint_hits, warning=FALSE}
table_endpoint_hits(chemicalSummary, category = "Biological")
# More options:
# table_endpoint_hits(chemicalSummary, category = "Chemical Class")
# table_endpoint_hits(chemicalSummary, category = "Chemical")
```

The tables show slightly different results for a single site. Instead of the number of sites with hits above a threshold, it is now the number of samples.

```{r table_endpoint_hits_site, warning=FALSE}
table_endpoint_hits(maumee, category = "Biological")
```


## table_tox_endpoint {#table_tox_endpoint}

```{r table_tox_endpoint, warning=FALSE}
table_tox_endpoint(chemicalSummary, category = "Chemical Class")
# More options:
# table_tox_endpoint(chemicalSummary, category = "Biological")
# table_tox_endpoint(chemicalSummary, category = "Chemical")
```

The tables show slightly different results for a single site, showing the number of samples with hits (instead of number of sites).

```{r table_tox_endpoint_site, warning=FALSE}
table_tox_endpoint(maumee, category = "Chemical Class")
```

# Maps {#make_tox_map}

There is one function included that creates a `leaflet` map of the data `make_tox_map`.

```{r makeMap, warning=FALSE, message=FALSE}
make_tox_map(chemicalSummary, tox_list$chem_site, "Biological")
# More options:
# make_tox_map(chemicalSummary, tox_list$chem_site, "Chemical Class")
# make_tox_map(chemicalSummary, tox_list$chem_site, "Chemical") 

```

