<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Introduction to toxEval}
-->

```{r setup, include=FALSE}
library(xtable)
options(continue=" ")
options(width=60)
library(knitr)
library(toxEval)
```

# Introduction to toxEval
By Laura A. De Cicco, Steven Corsi

This package **toxEval** is an early test version of the package that will encompass the various approaches to evaluate ToxCast data in relation to measured aquatic concentrations.

## Molecular weights
ToxCast reports values in $\mu$M. Data from water samples collected are typically mg/L, ng/L, or pg/L. Therefore, the first step is to determine the molecular weight of the measured data. The package **webchem** supports this functionality. 


```{r}

library("webchem")

inchk <- cts_convert(query = '3380-34-5', from = 'CAS', to = 'inchikey')
info <- cts_compinfo(inchikey = inchk)
info$molweight

```

## Workflow
Data measured in aquatic environments can come in many different formats. This package will use example data measured from passive samplers, as well as water samples taken at a variety of sites around the Great Lakes. 

### Passive Data:
Passive data was returned from the laboratory in a 'long' format. A sample of the data is shown below:

```{r echo=TRUE, eval=TRUE}

packagePath <- system.file("extdata", package="toxEval")
filePath <- file.path(packagePath, "passiveData.RData")
load(file=filePath)

head(passiveData[,1:5])

```


### Water Sample Data:
The data returned from water sample tests was returned in a 'wide' format:

```{r echo=TRUE, eval=TRUE}

packagePath <- system.file("extdata", package="toxEval")
filePath <- file.path(packagePath, "waterSamples.RData")
load(file=filePath)

head(waterSamples[,1:10])

```

