---
title: "Introduction to toxEval: Passive Data"
author: "Steven R. Corsi, Laura A. De Cicco"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    fig_caption: yes
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Introduction to toxEval}
  \usepackage[utf8]{inputenc}
---


```{r setup, include=FALSE}
options(continue=" ")
options(width=100)
library(knitr)
library(rmarkdown)
library(dplyr)
library(reshape2)
library(DT)

```

# Introduction

This package **toxEval** is an early test version of the package that will encompass the various approaches to evaluate ToxCast data in relation to measured aquatic concentrations.
<!--
## Molecular weights
ToxCast reports values in $\log$$\mu$M. Data from water samples collected are typically mg/L, ng/L, or pg/L. Therefore, the first step is to determine the molecular weight of the measured data. The package **webchem** supports this functionality. 


```{r}

library("webchem")

molweight <- cir_query('3380-34-5', "mw")
molweight

```

--> 
Data measured in aquatic environments can come in many different formats. This package will use example data measured from passive samplers, as well as water samples taken at a variety of sites around the Great Lakes. 

# Passive Data
Passive data was returned from the laboratory in a 'long' format. A sample of the data is shown below:

```{r echo=FALSE, eval=TRUE}

packagePath <- system.file("extdata", package="toxEval")
filePath <- file.path(packagePath, "passiveData.RData")
load(file=filePath)
#Site Info:
filePath <- file.path(packagePath, "stationINFO.RData")
load(file=filePath)
#parameter info:
filePath <- file.path(packagePath, "pCodeInfo.RData")
load(file=filePath)

#passive class info:
passiveCAS <- readRDS(file.path(packagePath, "passiveCAS.rds"))

pCodeInfo <- pCodeInfo[pCodeInfo$casrn != "", ]

# Some chemicals have 2 rows (different units)...
# To simplify for now, we'll ignore the duplicate chemicals:


passiveData <- left_join(passiveData, passiveCAS, by=c("CAS", "Chemical"))
passiveData <- passiveData[!duplicated(passiveData$CAS),]
passiveData$class[passiveData$Chemical == "Aspirin"] <-"Pharmaceuticals"

passiveData <- passiveData[,-which(names(passiveData) == "site04249000.1")]

head(passiveData[,1:7])

# Unique units:
unique(passiveData$Units)

```

For this data, the CAS column was used to retrieve the molecular weight (mlWt). Each column is a unique site where the data was collected. Each row is a unique measured chemical. 

Some of the data is 'left-censored'. This means that the measured data was lower than the detection limit of the measurement technique. Values that are found to be lower than the detection limit are indicated by a "<". The first thing to check is that the method quantitation limit (MQL) or minimum level of detectability (MLD) are low enough to be ignored. That is, MQL and MLD are below the minumum endpoint levels. If so, we can simply remove any values that have the "<" indicator. To begin the study of the endpoints and how they relate to our data, we need to load up the **toxEval** package, and choose an endpoint dataset. We will use the AC50gain dataset in this report.

For this workflow, we will use the units reported by the measured data, meaning we need to convert the ToxCast end points to the units in the "Units" column of the passiveData data frame (we could also do the analysis converting the measured data to $\log$$\mu$M... it shouldn't matter as long as you are consistent).

```{r message=FALSE}
library(toxEval)
endPointInfo <- endPointInfo
# AC50 data provided in the toxEval package:
AC50gain <- AC50gain
head(AC50gain[,1:5])

```

The units in the AC50gain, AC50loss, and AC10 data sets are $\log$$\mu$M. The first step is to convert to the desired unit (`r unique(passiveData$Units)`) based on the passiveData data frame.

$$
\require{cancel}
\begin{aligned}
Reported Data = \log \mu M \\
\mu M = 10^{\log \mu M} \\
MW = \frac{g}{mol} , M = \frac{mol}{L}  \\
\cancel{\mu M} \left[\frac{\cancel{M}}{10^6 \cancel{\mu M}}\right] \left[\frac{1}{\cancel{M}}  \frac{\cancel{mol}}{L}\right] \left[ \frac{MW \cancel{g}}{\cancel{mol}}\right] \left[\frac{10^9 ng}{\cancel{g}}\right] = MW * 10^3 \frac{ng}{L} 
\end{aligned}
$$

```{r message=FALSE}
library(dplyr)

unitConversion <- setNames(c(10^6, 10^3), c("pg/L", "ng/L") )

AC50 <- right_join(AC50gain[,c("casn"), drop=FALSE], 
                   passiveData[,c("CAS", "Units", "mlWt")],
                   by= c("casn" = "CAS")) %>%
  rename(desiredUnits = Units) %>%
  filter(!is.na(mlWt)) %>%
  mutate(conversion = unitConversion[desiredUnits] * mlWt) %>%
  select(casn, desiredUnits, mlWt, conversion)

```

A sample of the AC50 data frame is shown here:

```{r echo=FALSE}
kable(head(AC50), digits=2, row.names = FALSE)

```


Next we can create a data frame with the AC50 converted endpoints:
```{r}
AC50Converted <- left_join(AC50, AC50gain, by="casn")
infoColumns <- c("casn", "chnm", "desiredUnits","mlWt", "conversion", "code","chid")

endPointData <- AC50Converted[,!(names(AC50Converted) %in% infoColumns)]
endPointData <- 10^endPointData
endPointData <- endPointData * AC50Converted$conversion
endPoint <- cbind(AC50, data.frame(endPointData)) %>%
  rename(Units=desiredUnits) %>%
  filter(rowSums(is.na(endPointData)) != ncol(endPointData))

endPointData <- endPointData[rowSums(is.na(endPointData)) != 
                               ncol(endPointData),]

```



Now, let's look at the extreme reported endpoint (not including `r NA`'s):

```{r warning=FALSE}
maxEndPoints <- apply(endPointData, 1, max, na.rm=TRUE) 
minEndPoints <- apply(endPointData, 1, min, na.rm=TRUE)

maxMinSummary <- cbind(endPoint[,c("casn", "Units")], 
                       maxEndPoint=maxEndPoints, 
                       minEndPoint=minEndPoints) %>%
  arrange(desc(casn))

```

## Passive Detection Limits

Next, let's determine if the MQL or MLD are going to tell us anything in relation to the ToxCast endpoints. Likewise, we can look at the maximum and minimum measured values.

```{r warning=FALSE, echo=FALSE}
siteColumns <- grep("site",names(passiveData))
maxMinSummary <- left_join(maxMinSummary,passiveData[,-siteColumns], 
                           by=c("casn"="CAS", "Units"="Units")) 

```


If any of the MLD are higher than the maxEndPoint, we wouldn't be able to make any conclusions about the toxicity of the water.

```{r echo=FALSE}
sum(maxMinSummary$MLD > maxMinSummary$minEndPoint)

```

`r sum(maxMinSummary$MLD > maxMinSummary$maxEndPoint)` measurements have detection limitations that are greater than the minimum endpoint concentration. We might be interested to know how close the detection limit values come to the minimum endpoints. 

The following chemicals have MLD's within 1% of the miniumum endpoint value:

```{r echo=FALSE}
dfToPrint <- mutate(maxMinSummary, MLD_EAR = MLD/minEndPoint) %>%
  filter(MLD_EAR > 0.01) %>%
  select(Chemical, minEndPoint, MLD, MLD_EAR, class) %>%
  arrange(desc(MLD_EAR)) %>%
  rename("Max EAR"=MLD_EAR)
  
datatable(dfToPrint, rownames=FALSE) %>% 
    formatRound(c("Max EAR","MLD","minEndPoint"), digits = 3)

```

Since it appears that we can ignore the censored values, the easist way to convert data to numbers then is to use the `as.numeric` function. This will convert the left-censored data to `NA`, generating a warning. To convert all the columns with the word 'site' in their column name to numeric, use the following commands:

```{r warning=FALSE}
siteColumns <- grep("site",names(passiveData))
passiveData[,siteColumns] <- suppressWarnings(
  sapply(passiveData[,siteColumns], function(x) as.numeric(x)))

#For this analysis, we'll consider NA's to be 0 (other options exist):
passiveData[,siteColumns][is.na(passiveData[,siteColumns])] <- 0

```

## Chemical Summary Tables Passive

```{r message=FALSE, results='asis', echo=FALSE}
infoColumns <- c("Chemical", "CAS", "Units", "MLD", "MQL", "mlWt","class")

chemicalSummary <- melt(passiveData, id.vars = infoColumns) %>%
  mutate(variable=as.character(variable)) %>%
  filter(value != 0) %>%
  rename(measuredValue=value, site=variable) %>%
  right_join(endPoint, by=c("CAS"="casn", "Units"="Units", "mlWt"="mlWt")) %>%
  melt(id.vars = c(infoColumns, "site", "measuredValue","conversion")) %>% 
  mutate(variable=as.character(variable)) %>%
  rename(endPointValue=value, endPoint=variable) %>%
  filter(!is.na(endPointValue)) %>%
  mutate(EAR=measuredValue/endPointValue) %>%
  filter(EAR > 0.1) %>%
  group_by(Chemical, endPoint, class) %>%
  summarize(minEAR=min(EAR), maxEAR=max(EAR), hits=length(EAR), nSites=ncol(passiveData)) %>%
  left_join(endPointInfo, by=c("endPoint"="assay_component_endpoint_name")) %>%
  select(Chemical, minEAR, maxEAR, hits, nSites, endPoint, class, contains("intended_target_")) %>%
  arrange(Chemical, desc(maxEAR)) %>%
  rename(type=intended_target_type, type_sub=intended_target_type_sub,
         family=intended_target_family, family_sub=intended_target_family_sub) %>%
  select(-contains("intended_target_")) %>% 
    data.frame %>%
  arrange(desc(maxEAR)) 

  chemicalSummary <- unique(chemicalSummary) #some endpoints have multiple genes....so there are duplicated rows, there wouldn't be if we didn't have select(-contains("intended_target_"))

  datatable(chemicalSummary, rownames = FALSE, 
            options = list(pageLength = 10)) %>% 
    formatRound(c("minEAR","maxEAR"), digits = 2)

```

There are actually multiple genes for the endpoint that corresponds to Aspirin/`Tox21_TR_LUC_GH3_Antagonist`.

## Passive site summaries

```{r message=FALSE, results='asis', echo=FALSE}
infoColumns <- c("Chemical", "CAS", "Units", "MLD", "MQL", "mlWt","class")
siteKey <- setNames(stationINFO$shortName, stationINFO$STAID)

siteSummary <- melt(passiveData, id.vars = infoColumns) %>%
  mutate(variable=as.character(variable)) %>%
  filter(value != 0) %>%
  rename(measuredValue=value, site=variable) %>%
  right_join(endPoint, by=c("CAS"="casn", "Units"="Units", "mlWt"="mlWt")) %>%
  melt(id.vars = c(infoColumns, "site", "measuredValue","conversion")) %>% 
  select(Chemical, class, site, measuredValue, variable, value) %>%
  mutate(variable=as.character(variable)) %>%
  rename(endPointValue=value, endPoint=variable) %>%
  filter(!is.na(endPointValue)) %>%
  mutate(EAR=measuredValue/endPointValue) %>%
  filter(EAR > 0.1) %>%
  mutate(site=siteKey[gsub("site","",site)]) %>%
  group_by(site) %>%
  summarize(minEAR=min(EAR), maxEAR=max(EAR), hits=length(EAR), nChem=length(Chemical), nEndPoints=length(unique(endPoint))) %>%
  select(site, minEAR, maxEAR, hits, nChem, nEndPoints)  %>%
    data.frame %>%
  arrange(desc(maxEAR))  

  datatable(siteSummary, rownames = FALSE, 
            options = list(pageLength = 10)) %>% 
    formatRound(c("minEAR","maxEAR"), digits = 2)

```


## Passive Site tables
Let's create a set of summary tables for each site.

```{r warning=FALSE, echo=FALSE, results='asis'}
infoColumns <- c("Chemical", "CAS", "Units", "MLD", "MQL", "mlWt","class")

endpointNames <- names(endPoint)
endpointNames <- endpointNames[!(endpointNames %in% c("casn","Units","mlWt","conversion"))]

reversesiteKey <- setNames(paste0("site",stationINFO$STAID), stationINFO$shortName)

for(i in siteSummary$site){
  oneSite <- passiveData[,infoColumns]
  index <- reversesiteKey[i]
  oneSite$value <- passiveData[,index]
  
  oneSiteLong <- filter(oneSite, value != 0) %>%
    rename(measuredValue = value) %>%
#     left_join(pCodeInfo[c("parameter_cd","casrn","parameter_nm")], by=c("CAS"="casrn")) %>%
    select(Chemical, CAS, class, measuredValue) %>%
    right_join(endPoint, by=c("CAS"="casn")) %>%
    select(-mlWt, -conversion, -CAS,  -Units) %>%
    rename(chnm=Chemical) %>%
    melt(id.vars = c("measuredValue", "chnm", "class")) %>%
    mutate(variable=as.character(variable)) %>%
    rename(endPointValue=value, endPoint=variable) %>%
    filter(!is.na(endPointValue)) %>%
    mutate(EAR=measuredValue/endPointValue) %>%
    filter(EAR > 0.1) %>%
    left_join(endPointInfo, by=c("endPoint"="assay_component_endpoint_name")) %>%
    select(chnm, EAR, class, endPoint, contains("intended_target_")) %>%
    rename(type=intended_target_type, type_sub=intended_target_type_sub,
           family=intended_target_family, family_sub=intended_target_family_sub) %>%
    select(-contains("intended_target_")) %>%
    arrange(desc(EAR))
  
  oneSiteLong <- unique(oneSiteLong)
    

  if(nrow(oneSiteLong) != 0){
    cat("\n###",i)
    

#       datatable(oneSiteLong, rownames = FALSE) %>% formatRound(c("EAR"), digits = 2)
    print(kable(oneSiteLong, digits=2, caption = i, row.names = FALSE))
  }
  
  
}

```


## Passive visualizations

```{r fig.width=10, fig.height=10, warning=FALSE, echo=FALSE, fig.cap="Colors represent different endpoints, dots represent measured passive water sample values"}
library(ggplot2)
infoColumns <- c("Chemical", "CAS", "Units", "MLD", "MQL", "mlWt","class")

maxY <- 50000
dataPassive <- melt(passiveData, id.vars = infoColumns) %>%
  mutate(variable=as.character(variable)) %>%
  filter(value != 0) %>%
  rename(measuredValue=value, site=variable) %>%
  right_join(endPoint, by=c("CAS"="casn", "Units"="Units", "mlWt"="mlWt")) %>%
  melt(id.vars = c(infoColumns, "site", "measuredValue","conversion")) %>% 
  mutate(variable=as.character(variable)) %>%
  rename(endPointValue=value, endPoint=variable) %>%
  filter(!is.na(endPointValue)) %>%
  mutate(endPointValue=ifelse(endPointValue>maxY, maxY, endPointValue))

# ggplot(ep[1:50,], aes(variable)) + geom_bar() + facet_wrap(~ chnm)
# ggplot(ep[1:50,], aes(variable, fill=chnm)) + geom_bar() # + scale_fill_brewer()

ggplot(dataPassive[!is.na(dataPassive$measuredValue),], aes(x = Chemical, y = endPointValue, fill=endPoint)) +
    geom_bar(stat='identity',guide=FALSE,colour="grey") + 
    ylim(0,maxY) +   
    theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.25), legend.position = "none") +
    labs(x = "Chemical", y="Concentration [ng/L]") + 
    geom_point(aes(x=Chemical, y=measuredValue))

```

