<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Introduction to toxEval}
-->

```{r setup, include=FALSE}
library(xtable)
options(continue=" ")
options(width=100)
library(knitr)
```

# Introduction to toxEval
By Laura A. De Cicco, Steven Corsi

This package **toxEval** is an early test version of the package that will encompass the various approaches to evaluate ToxCast data in relation to measured aquatic concentrations.

## Molecular weights
ToxCast reports values in $\mu$M. Data from water samples collected are typically mg/L, ng/L, or pg/L. Therefore, the first step is to determine the molecular weight of the measured data. The package **webchem** supports this functionality. 


```{r}

library("webchem")

inchk <- cts_convert(query = '3380-34-5', from = 'CAS', to = 'inchikey')
info <- cts_compinfo(inchikey = inchk)
info$molweight

```

## Workflow
Data measured in aquatic environments can come in many different formats. This package will use example data measured from passive samplers, as well as water samples taken at a variety of sites around the Great Lakes. 

### Passive Data:
Passive data was returned from the laboratory in a 'long' format. A sample of the data is shown below:

```{r echo=TRUE, eval=TRUE}

packagePath <- system.file("extdata", package="toxEval")
filePath <- file.path(packagePath, "passiveData.RData")
load(file=filePath)

head(passiveData[,1:8])

# Unique units:
unique(passiveData$Units)

```

For this data, the CAS column was used to retrieve the molecular weight (mlWt). Each column is a unique site where the data was collected. Each row is a unique measured chemical. The data may be 'left-censored'. This means that the measured data was lower than the detection limit of the measurement technique. Values that are found to be lower than the detection limit are indicated by a "<".

The first task is to determine if the method quantitation limit (MQL) or minimum level of detectability (MLD) are going to tell us anything in relation to the ToxCast endpoints. We have to choose whether or not to move forward with this investigation in units reported by the measured data, or units of ToxCast endpoints. For this workflow, we will use the units reported by the measured data, meaning we need to convert the ToxCast end points to the units in the "Units" column in the passiveData data frame.


```{r message=FALSE}
library(toxEval)

# AC50 data provided in the toxEval package:
AC50gain <- AC50gain

head(AC50gain[,1:5])

```

The units in the AC50gain, AC50loss, and AC10 data sets are in log base 10 $\mu$M. The first step is to convert to the desired unit (`r unique(passiveData$Units)`)

First, let's merge the desired units to the data frame:

$$
\begin{aligned}
mlWt = \frac{g}{mol} \\
M = \frac{mol}{L} \\
\frac{\mu M}{} \frac{M}{10^6 \mu M} \frac{1}{M}  \frac{mol}{L}  \frac{1}{mlWt}  \frac{g}{mol} \frac{10^9 ng}{g} = \frac{10^3}{mlWt} \frac{ng}{L}
\end{aligned}
$$

```{r message=FALSE}
library(dplyr)

unitConversion <- setNames(c(1E-12, 1E-9), c("pg/L", "ng/L") )

AC50 <- left_join(AC50gain, passiveData[,c("CAS", "Units", "mlWt")],
                   by= c("casn" = "CAS")) %>%
  select(casn, chnm, Units, mlWt) %>%
  filter(!is.na(Units)) 

```




Let's start our analysis with one site (the first of the site columns: site04101500):

```{r, message=FALSE}
commonColumns <- c("Chemical", "Units", 
                  "MLD", "MQL", "CAS", "mlWt")
oneSite <- passiveData[,c(commonColumns, "site04101500")]


oneSite <- rename(oneSite, value = site04101500)
head(oneSite)

```


Because of the "<", the measured data is currently a vector of characters. The simpliest way to convert that data to numbers is to use the `as.numeric` function. This will convert the left-censored data to `NA`, generating a warning. We can discuss more vigorious methods for dealing with censored values in future sections.

```{r}
oneSite$value <- as.numeric(oneSite$value)

head(oneSite$value)

```


### Water Sample Data:
The data returned from water sample tests was returned in a 'wide' format:

```{r echo=TRUE, eval=TRUE}

packagePath <- system.file("extdata", package="toxEval")
filePath <- file.path(packagePath, "waterSamples.RData")
load(file=filePath)

head(waterSamples[,1:10])

```

