<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Introduction to toxEval}
-->

```{r setup, include=FALSE}
library(xtable)
options(continue=" ")
options(width=100)
library(knitr)
```

# Introduction to toxEval
By Laura A. De Cicco, Steven Corsi

This package **toxEval** is an early test version of the package that will encompass the various approaches to evaluate ToxCast data in relation to measured aquatic concentrations.

## Molecular weights
ToxCast reports values in $\log$$\mu$M. Data from water samples collected are typically mg/L, ng/L, or pg/L. Therefore, the first step is to determine the molecular weight of the measured data. The package **webchem** supports this functionality. 


```{r}

library("webchem")

inchk <- cts_convert(query = '3380-34-5', from = 'CAS', to = 'inchikey')
info <- cts_compinfo(inchikey = inchk)
info$molweight

```

## Workflow
Data measured in aquatic environments can come in many different formats. This package will use example data measured from passive samplers, as well as water samples taken at a variety of sites around the Great Lakes. 

### Passive Data:
Passive data was returned from the laboratory in a 'long' format. A sample of the data is shown below:

```{r echo=TRUE, eval=TRUE}

packagePath <- system.file("extdata", package="toxEval")
filePath <- file.path(packagePath, "passiveData.RData")
load(file=filePath)

head(passiveData[,1:8])

# Unique units:
unique(passiveData$Units)

```

For this data, the CAS column was used to retrieve the molecular weight (mlWt). Each column is a unique site where the data was collected. Each row is a unique measured chemical. The data may be 'left-censored'. This means that the measured data was lower than the detection limit of the measurement technique. Values that are found to be lower than the detection limit are indicated by a "<".

Because of the "<", the measured data is currently a vector of characters. The simpliest way to convert that data to numbers is to use the `as.numeric` function. This will convert the left-censored data to `NA`, generating a warning. We can discuss more vigorious methods for dealing with censored values in future sections. To convert all the columns with the word 'site' in their column name, use the following commands:

```{r warning=FALSE}
siteColumns <- grep("site",names(passiveData))
passiveData[,siteColumns] <- sapply(passiveData[,siteColumns], function(x) as.numeric(x))

# #Let's remove rows that now have all NA's:
# passiveData <- passiveData[rowSums(is.na(passiveData[,siteColumns]))!=length(siteColumns) ,]

```


For this workflow, we will use the units reported by the measured data, meaning we need to convert the ToxCast end points to the units in the "Units" column of the passiveData data frame (we could also do the analysis converting the measured data to $\log$$\mu$M... it shouldn't matter as long as you are consistent).


```{r message=FALSE}
library(toxEval)

# AC50 data provided in the toxEval package:
AC50gain <- AC50gain

head(AC50gain[,1:5])

```

The units in the AC50gain, AC50loss, and AC10 data sets are in $\log$$\mu$M. The first step is to convert to the desired unit (`r unique(passiveData$Units)`)

First, let's merge the desired units to the data frame:

$$
\require{cancel}
\begin{aligned}
MW = \frac{g}{mol} \\
M = \frac{mol}{L} \\
\cancel{\mu M} \left[\frac{\cancel{M}}{10^6 \cancel{\mu M}}\right] \left[\frac{1}{\cancel{M}}  \frac{\cancel{mol}}{L}\right] \left[ \frac{MW \cancel{g}}{\cancel{mol}}\right] \left[\frac{10^9 ng}{\cancel{g}}\right] = MW * 10^3 \frac{ng}{L} 
\end{aligned}
$$

```{r message=FALSE}
library(dplyr)

unitConversion <- setNames(c(10^6, 10^3), c("pg/L", "ng/L") )

AC50 <- left_join(AC50gain, passiveData[,c("CAS", "Units", "mlWt")],
                   by= c("casn" = "CAS")) %>%
  filter(!is.na(Units)) %>%
  rename(desiredUnits = Units) %>%
  mutate(conversion = unitConversion[desiredUnits] * mlWt) %>%
  select(casn, chnm, desiredUnits, mlWt, conversion)

```

A sample of the AC50 data frame is shown here:

```{r echo=FALSE}
kable(head(AC50))

```

Next we can create a data frame with the AC50 converted endpoints:
```{r}
AC50Converted <- left_join(AC50, AC50gain)
infoColumns <- c("casn", "chnm", "desiredUnits","mlWt", "conversion", "code","chid")

endPointData <- AC50Converted[,!(names(AC50Converted) %in% infoColumns)]
endPointData <- 10^endPointData
endPointData <- endPointData * AC50Converted$conversion
endPoint <- cbind(AC50, data.frame(endPointData))
endPoint <- rename(endPoint, Units=desiredUnits)

```

First, let's look at the extreme reported endpoint (not including `r NA`'s...that will actually be very important):

```{r warning=FALSE}
maxEndPoints <- apply(endPointData, 1, max, na.rm=TRUE) 
minEndPoints <- apply(endPointData, 1, min, na.rm=TRUE)

maxMinSummary <- cbind(endPoint[,c("chnm", "casn", "Units")], 
                       maxEndPoint=maxEndPoints, 
                       minEndPoint=minEndPoints)

```

The next task is to determine if the method quantitation limit (MQL) or minimum level of detectability (MLD) are going to tell us anything in relation to the ToxCast endpoints. Likewise, we can look at the maximum and minimum measured values.

```{r warning=FALSE}

maxMeasure <- apply(passiveData[,siteColumns], 1, max, na.rm=TRUE) 
minMeasure <- apply(passiveData[,siteColumns], 1, min, na.rm=TRUE)
dataSummary <- passiveData[,-siteColumns]
dataSummary$Max <- maxMeasure
dataSummary$Min <- minMeasure

maxMinSummary <- left_join(maxMinSummary,dataSummary, 
                           by=c("casn"="CAS", "Units"="Units"))

```

A sample of the maxMinSummary data frame is shown here: 

```{r echo=FALSE}
kable(head(maxMinSummary[,c(1:5,7:8,10:11)]), digits=3)

```

If any of the MLD are higher than the maxEndPoint, we wouldn't be able to make any conclusions about the toxicity of the water:

```{r}
sum(maxMinSummary$MLD > maxMinSummary$maxEndPoint)

```


### Data Analysis
Let's start our analysis with one site (the first of the site columns: site04101500):

```{r message=FALSE}
commonColumns <- c("Chemical", "Units", 
                  "MLD", "MQL", "CAS", "mlWt")
oneSite <- passiveData[,c(commonColumns, "site04101500")]


oneSite <- rename(oneSite, value = site04101500)
head(oneSite)

```



### Water Sample Data:
The data returned from water sample tests was returned in a 'wide' format:

```{r echo=TRUE, eval=TRUE}

packagePath <- system.file("extdata", package="toxEval")
filePath <- file.path(packagePath, "waterSamples.RData")
load(file=filePath)

head(waterSamples[,1:10])

```

