---
title: "Setting up toxEval package data"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: false
vignette: >
  %\VignetteIndexEntry{Setting up toxEval package data}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = FALSE,
                      warning = FALSE,
                      message = FALSE)
library(tcpl)
```

`toxEval` includes ToxCast data to help simplify EAR calculations. Here are the instructions for getting the necessary columns from the ToxCast database using the `tcpl` package. This is NOT necessary for all `toxEval` users. However, if you are interested in extending the data or exploring additional fields, this might be useful. It is also a way to remember how to update the `toxEval` package when there are new ToxCast database updates.

First you would need to download the database and set up a MySQL server/database. This is beyond the scope of this vigentte right now. 

Next, you need to connect to that local database:

```{r}
library(tcpl)
library(toxEval)
library(tidyverse)


tcplConf(drvr = "MySQL", 
         user = "root",
         pass = "my_super_secret_password",
         host = "localhost", 
         db = "prod_internal_invitrodb_v4_1")


```

## ToxCast_ACC

First, we'll need to load several tables to get our necessary columns:

```{r}
mc5_raw <- tcplQuery("SELECT   * FROM mc5")
mc4_raw <- tcplQuery("SELECT   * FROM mc4")
mc5_chid <- tcplQuery("SELECT   * FROM mc5_chid")
mc6_raw <- tcplQuery("SELECT   * FROM mc6")
sample_raw <- tcplQuery("SELECT   * FROM sample")
chemical_raw <- tcplQuery("SELECT   * FROM chemical")
mc5_param <- tcplQuery("SELECT   * FROM mc5_param")
```

Next, we'll use `dplyr` to join:

```{r eval=FALSE}
ToxCast_ACC <- mc5_raw |> 
  filter(hitc >= 0.9) |> 
  select(-created_date, -modified_date, -modified_by) |> 
  left_join(mc4_raw |> 
              select(-created_date, -modified_date, -modified_by),
            by = join_by(m4id, aeid)) |> 
  left_join(mc5_chid, by = join_by(m5id)) |> 
  filter(chid_rep == 1) |> 
  left_join(mc6_raw |> 
              select(-created_date, -modified_date, -modified_by),
            by = join_by(m5id, m4id, aeid)) |> 
  left_join(sample_raw,
            by = join_by(spid)) |> 
  left_join(chemical_raw,
            by = join_by(chid)) |> 
  left_join(mc5_param |> 
              filter(hit_param == "acc"),
            by = join_by(m5id, aeid)) |> 
  select(casn, hit_val, aeid, mc6_mthd_id) |> 
  group_by(casn, hit_val, aeid) |> # 1-to-many flags
  summarise(flags = paste(mc6_mthd_id, collapse = ", ")) |> 
  ungroup()

names(ToxCast_ACC)

```

```
[1] "casn"    "hit_val" "aeid"    "flags" 
```

## end_point_info

Next, we'll populate the `end_point_info` file:

```{r}

assay_component_endpoint <- tcplQuery("SELECT * FROM assay_component_endpoint;")
assay <- tcplQuery("SELECT * FROM assay;")
assay_component <- tcplQuery("SELECT * FROM assay_component;")
assay_source <- tcplQuery("SELECT * FROM assay_source;")
gene <- tcplQuery("SELECT * FROM gene;")
intended_target <- tcplQuery("SELECT * FROM intended_target;")

end_point_info_41 <- assay_component_endpoint |> 
  left_join(assay_component, by = join_by(acid)) |> 
  left_join(assay, by = join_by(aid)) |> 
  left_join(assay_source, by = join_by(asid)) |> 
  left_join(intended_target, by = join_by(aeid)) |> 
  left_join(gene, by = c("target_id" = "gene_id")) |> 
  filter(!is.na(aeid)) |> 
  select(aeid, acid, assay_component_endpoint_name, 
         assay_component_endpoint_desc, intended_target_type, 
         intended_target_family_sub, intended_target_family,
         biological_process_target, tissue, 
         gene_symbol, assay_source_name) |> 
  group_by(across(c(-gene_symbol))) |> # 1-to-many genes
  summarise(gene_symbol = paste(gene_symbol, collapse = ", ")) |> 
  ungroup()

```

## tox_chemicals

```{r}
tox_chemicals41 <- mc5_raw |> 
  select(-created_date, -modified_date, -modified_by) |> 
  left_join(mc4_raw |> 
              select(-created_date, -modified_date, -modified_by),
            by = join_by(m4id, aeid)) |> 
  left_join(mc5_chid, by = join_by(m5id)) |> 
  filter(chid_rep == 1) |> 
  left_join(mc6_raw |> 
              select(-created_date, -modified_date, -modified_by),
            by = join_by(m5id, m4id, aeid)) |> 
  left_join(sample_raw,
            by = join_by(spid)) |> 
  left_join(chemical_raw,
            by = join_by(chid)) |> 
  left_join(mc5_param |> 
              filter(hit_param == "acc"),
            by = join_by(m5id, aeid)) |> 
  group_by(casn, chnm, dsstox_substance_id) |> 
  summarise(Total_tested = length(unique(aeid)),
            Active = length(unique(aeid[hitc >= 0.9]))) |> 
  ungroup() |> 
  left_join(tox_chemicals_35 |> 
              select(casn = Substance_CASRN,
                     Structure_MolWt),
            by = "casn")
```

We need molecular weights to convert the ACC values to concentrations. We can either join previous versions of the toxEval package, and/or get new values from the CompTox Dashboard via the batch search:

<https://comptox.epa.gov/dashboard/batch-search\>

To create a list of chemicals to input to the dashboard:

```{r}
need_mlwt <- tox_chemicals41$dsstox_substance_id[is.na(tox_chemicals41$Structure_MolWt)]
#Bring this to Comptox:
data.table::fwrite(data.frame(need_mlwt), "need_mlwt.csv")  
```


```{r}
# Import the file from Comptox:
more_mlwts <- data.table::fread("CCD-Batch-Search.csv",
                                data.table = FALSE) |> 
  mutate(new_Structure_MolWt = as.numeric(AVERAGE_MASS)) |> 
  select(casn = CASRN,
         dsstox_substance_id = DTXSID,
         new_Structure_MolWt)

tox_chemicals41 <- tox_chemicals41 |> 
  left_join(more_mlwts, 
              by = c("casn", "dsstox_substance_id")) |> 
  mutate(Structure_MolWt = if_else(is.na(Structure_MolWt),
                                        new_Structure_MolWt,
                                        Structure_MolWt)) |> 
  select(-new_Structure_MolWt)
```


## sysdata.rda

Finally we can save that data in the package:

```{r}
ToxCast_ACC <- ToxCast_ACC_41
tox_chemicals <- tox_chemicals41
end_point_info <- end_point_info_41_rel

save(ToxCast_ACC, tox_chemicals, end_point_info,
     file = "sysdata.rda", compress = "xz")

```
