<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Introduction to toxEval}
-->

```{r setup, include=FALSE}
library(xtable)
options(continue=" ")
options(width=60)
library(knitr)

```

# Introduction to toxEval
By Laura A. De Cicco, Steven Corsi

This package **toxEval** is an early test version of the package that will encompass the various approaches to evaluate ToxCast data in relation to measured aquatic concentrations.

## Molecular weights
ToxCast reports values in $\mu$M. Data from water samples collected are typically mg/L, ng/L, or pg/L. Therefore, the first step is to determine the molecular weight of the measured data. The package **webchem** supports this functionality. 


```{r}

library("webchem")

inchk <- cts_convert(query = '3380-34-5', from = 'CAS', to = 'inchikey')
info <- cts_compinfo(inchikey = inchk)
info$molweight

```

## Workflow
Data measured in aquatic environments can come in many different formats. This package will use example data measured from passive samplers, as well as water samples taken at a variety of sites around the Great Lakes. 


```{r}
library(toxEval)
filePath <- system.file("extdata", package="toxEval")
passiveData <- load(file.path(filePath, "passiveData.RData"))

head(passiveData)

```


